---
title: "WHO data"
output:
  html_notebook: default
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 2
---

```{r}
library(gridExtra)
library(ggpubr)
library(stringr)
library(dplyr)
```

# Why: Why is this a real problem that is worth solving?

We perform exploratory analysis on World Health Organization (WHO) datasets.


```{r}
mortality_rate <- read.csv(
  "mortality_rate.csv",
  stringsAsFactors = F
) 

mortality_rate <- mortality_rate %>%  
  mutate(
    Number.of.under.five.deaths..thousands. = as.numeric(Number.of.under.five.deaths..thousands.),
    Number.of.infant.deaths..thousands. = as.numeric(Number.of.infant.deaths..thousands.),
    Number.of.neonatal.deaths..thousands. = as.numeric(Number.of.neonatal.deaths..thousands.)
  )

mortality_rate[is.na(mortality_rate)] <- 0
```

```{r}
healthcare_expenditure <- read.csv(
  "healthcare_expenditure.csv",
  stringsAsFactors = F
)

healthcare_gdp <- bind_rows(
  healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP.....,
      Year = 2015
    ),
  healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......1,
      Year = 2014
    ),
  healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......2,
      Year = 2013
    ),
  healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......3,
      Year = 2012
    ),
  healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......4,
      Year = 2011
    ),
  healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......5,
      Year = 2010
    ),
    healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......6,
      Year = 2009
    ),
    healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......7,
      Year = 2008
    ),
    healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......8,
      Year = 2007
    ),
    healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......9,
      Year = 2006
    ),
    healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......10,
      Year = 2005
    ),
    healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......11,
      Year = 2004
    ),
      healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......12,
      Year = 2003
    ),
      healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......13,
      Year = 2002
    ),
      healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......14,
      Year = 2001
    ),
      healthcare_expenditure %>% 
    transmute(
      Country = X,
      healthcare_expenditure_perc_gdp = Current.health.expenditure..CHE..as.percentage.of.gross.domestic.product..GDP......15,
      Year = 2000
    )
)

healthcare_gdp <- healthcare_gdp %>% 
  mutate(
    healthcare_expenditure_perc_gdp = as.numeric(healthcare_expenditure_perc_gdp)
  )

```


### How does mortality rate compare for so called developing countries against United States?

```{r}
mortality_rate %>% 
  filter(Country %in% c('Mongolia', 'United States of America', 'Mexico', 'Sweden')) %>% 
  ggplot(aes(x = Year, y = Number.of.under.five.deaths..thousands., color = Country)) + 
    geom_line()
```

As expected, Sweden has a very good mortality rate, but Mongolia has made tremendous progress towards reducing the mortality rate (which is now comparable to that of Sweden). It is very odd that US has extremely high mortality rate. 

### How does current mortality rate looks among countries?

```{r}
merge(
  mortality_rate,
  healthcare_gdp,
  by=c('Country', 'Year')
) %>% 
  filter(
    Year == 2015,
    Country %in% c('Mongolia', 'United States of America', 'Mexico', 'Sweden')
  ) %>% 
  ggplot(aes(x = healthcare_expenditure_perc_gdp, y = Number.of.under.five.deaths..thousands., label = Country)) + 
    geom_point() + 
    geom_text()
```

How is the correlation between healthcare expenditure percenatage and mortality rate for all countries as of 2015?

```{r}
merge(
  mortality_rate,
  healthcare_gdp,
  by=c('Country', 'Year')
) %>% 
  filter(
    Year == 2015
  ) %>% 
  ggplot(aes(x = healthcare_expenditure_perc_gdp, y = Number.of.under.five.deaths..thousands., label = Country)) + 
    geom_point() + 
    geom_text()
```



# How: How do we know if we have solved the problem?

TODO: Explain the process of addressing problems 

TODO: Include metrics we will use to define our success

# What: What does this look like in the product?

TODO: Summarize the takeaways

TODO: Provide details on the solution and productionalization plan


# Timeline

1. Finish 1-pager [2018-01-22]
2. Final deliverable [2018-02-29]

# References

https://www.who.int/gho/en/