---
title: "presidential_election"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 2
    code_folding: show
  html_notebook: default
---

```{r include=FALSE}
library(tidyverse)
library(ggrepel)
```

# Why: Why is this a real problem that is worth solving?

In this document, we perform basic exploraty analysis to gain some insights into US Presidential election. 


Source: https://dataverse.harvard.edu/file.xhtml?persistentId=doi:10.7910/DVN/42MVDX/MFU99O&version=5.0

```{r}
full_data <- read_csv("dataset/1976-2016-president.csv") 

# for the most of the analysis, we will keep our focus in 2 of the biggest parties
DEM <- "democrat"
REP <- "republican"
big2_data <- full_data %>% 
  filter(party %in% c(DEM, REP)) %>% 
  mutate(candidatevotes_in_millions = candidatevotes / 1e6)
```



### Total votes across all states by party

```{r}
big2_data %>% 
  group_by(candidate, party, year) %>% 
  summarise(total_vote = sum(candidatevotes_in_millions)) %>% 
  filter(total_vote > 5) %>% 
  ggplot(aes(x = year, y = total_vote, color = party)) + 
    geom_line() + 
    geom_point() + 
    geom_text(aes(label = candidate), vjust = -0.5, check_overlap = T, size=3) +
    ylab("Total vote in Millions")
```

NOTE: For the last 3 elections, democratis have won the popular vote.

### Democratic candidate votes in certain states of interests

```{r}
state_plot <- function(st) {
  state_name <- big2_data %>% filter(state_po == st) %>% select(state) %>% unique()
  big2_data %>% 
    filter(state_po == st) %>%  
    ggplot(aes(x = year, y = candidatevotes_in_millions, color = party)) + 
      geom_line() + 
      geom_point() + 
      labs(title=paste0("Party total vote comparison in ", state_name)) 
}

state_plot("CA")
```

NOTES: 
1. Interestingly around 30 years ago, CA was actually a republican state, and since then it's been mostly DEM and ever increasing

```{r}
state_plot("PA")
```

```{r}
state_plot("WI")
```

```{r}
state_plot("FL")
```


### Popularity margin of Democratic candidate

```{r}
big2_data %>% 
  group_by(year, party) %>% 
  summarise(
    candidate_total = sum(candidatevotes) / sum(totalvotes)
  ) %>% 
  pivot_wider(names_from = party, values_from = candidate_total) %>% 
  mutate(democrat_winning_margin = democrat - republican) %>% 
  ggplot(aes(x = year, y = democrat_winning_margin)) + 
    geom_point() + 
    geom_line()
```

NOTE: It is very much cyclical


### Who has won by biggest margin?

```{r}
widest_margin_year <- big2_data %>% 
  group_by(year, party) %>% 
  summarise(
    candidate_total = sum(candidatevotes) / sum(totalvotes)
  ) %>% 
  pivot_wider(names_from = party, values_from = candidate_total) %>% 
  mutate(margin = abs(democrat - republican)) %>% 
  arrange(desc(margin)) %>% 
  head(1)

widest_margin_year %>% 
  inner_join(big2_data, by="year") %>% 
  group_by(candidate) %>% 
  summarise(total_votes_in_millions = sum(candidatevotes_in_millions)) %>% 
  ggplot(aes(x = candidate, y = total_votes_in_millions, fill = candidate)) + 
    geom_bar(stat="identity") + 
    geom_text(aes(label = round(total_votes_in_millions, 2)), vjust = -0.5) + 
    labs(title = paste0(
      "Highest margin between 2 parties was in ", 
      widest_margin_year %>% select(year)
    ))
```

# Summary

* For last 3 elections, Democratic party has won the popular vote but the margin has been decreasing
* We observe ever growing margin of democratic turn for California
* So called swing states of PA, Wisconsin, Florida all have elected Republican for the 2016 election (but all has elected Obama for 2008 and 2012)
* Ronald Reagon was the single most popular candidate who won by more than 16% margin in 1984.